---
title: "Clark EDA"
author: "Dan Clark"
format: pdf
---

Claude Opus 5 was used to brainstorm ways to turn EDA findings into charts that would show a lot of information in a limited number of pages. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, autodep = TRUE,
  fig.align = "center")
```

```{r}
#| warning: FALSE
#| message: FALSE
library(tidyverse)
library(tidymodels)
library(here)
library(patchwork)
```

```{r}
set.seed(42)

earthquake <- readRDS(here("data", "processed", "earthquake.rds"))
#dim(earthquake)
#head(earthquake)
#glimpse(earthquake)
```

```{r}
colSums(is.na(earthquake))
```

```{r}
earthquake <- earthquake |>
  mutate(damage_grade = factor(damage_grade, levels = c(1, 2, 3)))

earthquake_split <- initial_split(earthquake, prop = 0.80, strata = damage_grade)

train_data <- training(earthquake_split)
holdout_data <- testing(earthquake_split)
```

### Basic Data Checks

- No missing values were found in any variable.
- The dataset contains a mix of:
  - Numeric building characteristics such as `age`, `area_percentage`, and `height_percentage`
  - Categorical structural variables such as `foundation_type`, `roof_type`, and `plan_configuration`
  - Binary indicator variables for superstructure and secondary-use characteristics
  - Three geographic ID variables at different levels of granularity
- `damage_grade` was converted from numeric values to a categorical factor
- The three geographic ID variables are stored numerically but represent geographic categories rather than continuous measurements.
- The data were split into 80% test and 20% holdout sets using stratification by `damage_grade`.

```{r}
damage_dist <- train_data |>
  count(damage_grade) |>
  mutate(percent = 100 * n / sum(n))

ggplot(damage_dist, aes(x = damage_grade, y = percent)) +
  geom_col() +
  geom_text(aes(label = paste0(round(percent, 1), "%")), vjust = -0.3) +
  labs(title = "Distribution of Damage Grade", x = "Damage Grade", y = "Percent of Training Set") +
  ylim(0, 65)
```

- Damage grade is imbalanced.
  - Grade 2 is 56.9% of the training set.
  - Grade 3 is 33.4%.
  - Grade 1 is 9.7%.

```{r}
train_data |>
  filter(age < 995) |>
  ggplot(aes(x = damage_grade, y = age)) +
  geom_boxplot() +
  labs(title = "Building Age by Damage Grade", subtitle = "Age = 995 excluded from visualization only", x = "Damage Grade", y = "Building Age")
```

- Building age shows a relationships across damage grade and a large number of outliers. 
  -There is a large outlier at 995 with 1,123 observations in the training set. Removed it from the chart so it works visually but we will revisit it in each model.

```{r}
#| fig-width: 7
#| fig-height: 8
#| out-width: "100%"

geo1_plot <- train_data |>
  count(geo_level_1_id, damage_grade) |>
  group_by(geo_level_1_id) |>
  mutate(percent = 100 * n / sum(n)) |>
  ungroup()

geo1_order <- geo1_plot |>
  filter(damage_grade == "3") |>
  arrange(percent) |>
  pull(geo_level_1_id)

geo1_plot <- geo1_plot |>
  mutate(geo_level_1_id = factor(geo_level_1_id, levels = geo1_order))

ggplot(geo1_plot, aes(x = geo_level_1_id, y = percent, fill = damage_grade)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(
      "1" = "light green",
      "2" = "orange",
      "3" = "red")) +
  labs(title = "Damage Grade Distribution by Geographic Level 1", x = "Geographic Level 1 ID", y = "Percent of Buildings", fill = "Damage Grade")
```

```{r}
#| fig-width: 7
#| fig-height: 8
#| out-width: "100%"

geo2_largest <- train_data |>
  count(geo_level_2_id) |>
  slice_max(n, n = 30) |>
  pull(geo_level_2_id)

geo2_plot <- train_data |>
  filter(geo_level_2_id %in% geo2_largest) |>
  count(geo_level_2_id, damage_grade) |>
  group_by(geo_level_2_id) |>
  mutate(percent = 100 * n / sum(n)) |>
  ungroup()

geo2_order <- geo2_plot |>
  filter(damage_grade == "3") |>
  arrange(percent) |>
  pull(geo_level_2_id)

geo2_plot <- geo2_plot |>
  mutate(geo_level_2_id = factor(geo_level_2_id, levels = geo2_order))

ggplot(geo2_plot, aes(x = geo_level_2_id, y = percent, fill = damage_grade)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(
      "1" = "light green",
      "2" = "orange",
      "3" = "red")) +
  labs(title = "Damage Grade Distribution by Geographic Level 2",
       x = "Geographic Level 2 ID",
       y = "Percent of Buildings",
       fill = "Damage Grade")
```

```{r}
#| fig-width: 7
#| fig-height: 8
#| out-width: "100%"

geo3_largest <- train_data |>
  count(geo_level_3_id) |>
  slice_max(n, n = 30) |>
  pull(geo_level_3_id)

geo3_plot <- train_data |>
  filter(geo_level_3_id %in% geo3_largest) |>
  count(geo_level_3_id, damage_grade) |>
  group_by(geo_level_3_id) |>
  mutate(percent = 100 * n / sum(n)) |>
  ungroup()

geo3_order <- geo3_plot |>
  filter(damage_grade == "3") |>
  arrange(percent) |>
  pull(geo_level_3_id)

geo3_plot <- geo3_plot |>
  mutate(geo_level_3_id = factor(geo_level_3_id, levels = geo3_order))

ggplot(geo3_plot, aes(x = geo_level_3_id, y = percent, fill = damage_grade)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(
  "1" = "#9ECAE1",
  "2" = "#4292C6",
  "3" = "#084594"
)) +
  labs(title = "Damage Grade Distribution by Geographic Level 3",
       x = "Geographic Level 3 ID",
       y = "Percent of Buildings",
       fill = "Damage Grade")
```


```{r}
selected_superstructures <- c(
  "has_superstructure_rc_engineered",
  "has_superstructure_rc_non_engineered",
  "has_superstructure_cement_mortar_brick",
  "has_superstructure_cement_mortar_stone",
  "has_superstructure_mud_mortar_stone",
  "has_superstructure_stone_flag"
)

superstructure_plot <- train_data |>
  select(damage_grade,all_of(selected_superstructures)) |>
  pivot_longer(cols = -damage_grade, names_to = "variable", values_to = "has_feature") |>
  mutate(has_feature = factor(has_feature, levels = c(0, 1), labels = c("No", "Yes")), variable = recode(variable,
      "has_superstructure_rc_engineered" = "RC engineered",
      "has_superstructure_rc_non_engineered" = "RC non-engineered",
      "has_superstructure_cement_mortar_brick" = "Cement mortar brick",
      "has_superstructure_cement_mortar_stone" = "Cement mortar stone",
      "has_superstructure_mud_mortar_stone" = "Mud mortar stone",
      "has_superstructure_stone_flag" = "Stone flag")) |>
  count(variable, has_feature, damage_grade) |>
  group_by(variable, has_feature) |>
  mutate(percent = 100 * n / sum(n)) |>
  ungroup()

ggplot(superstructure_plot, aes(x = has_feature, y = percent, fill = damage_grade)) +
  geom_col() +
  facet_wrap(~ variable, ncol = 3) +
  scale_fill_manual(values = c(
  "1" = "#9ECAE1",
  "2" = "#4292C6",
  "3" = "#084594"
)) +
  labs(title = "Damage Grade by Selected Superstructure Features", y = "Percent of Buildings", fill = "Damage Grade")
```

```{r}
selected_structural <- c(
  "foundation_type",
  "roof_type",
  "ground_floor_type",
  "other_floor_type"
)

structural_plot <- train_data |>
  select(damage_grade, all_of(selected_structural)) |>
  pivot_longer(cols = -damage_grade, names_to = "variable", values_to = "level") |>
  mutate(variable = recode(variable,
      "foundation_type" = "Foundation type",
      "roof_type" = "Roof type",
      "ground_floor_type" = "Ground floor type",
      "other_floor_type" = "Other floor type")) |>
  count(variable, level, damage_grade) |>
  group_by(variable, level) |>
  mutate(percent = 100 * n / sum(n)) |>
  ungroup()

ggplot(structural_plot,
  aes(x = level, y = percent, fill = damage_grade)) +
  geom_col() +
  facet_wrap(~ variable, ncol = 2, scales = "free_x") +
  scale_fill_manual(values = c(
  "1" = "#9ECAE1",
  "2" = "#4292C6",
  "3" = "#084594"
)) +
  labs(title = "Damage Grade by Selected Structural Characteristics", x = "Category", y = "Percent of Buildings", fill = "Damage Grade")
```

## Feature Selection

- Use the EDA and model validation to guide feature selection
- Remove `building_id` and `has_secondary_use`
- Use cross-validation to determine if predictors improve performance or not.
- Geography showed a strong relationship with damage severity.
- Review predictors with very rare or sparse categories.
  - `plan_configuration` levels, several detailed secondary-use categories, and sparse `geo_level_3_id` regions.

## Model Selection

One of the models we plan to use is XGBoost. XGBoost is a good fit for this dataset because the EDA suggests nonlinear relationships and potentially important interactions among geographic, structural, and building characteristics. XGBoost is also a tunable model, which allows model complexity to be selected through cross-validation.  

## Training and Validation

- Use 80/20 stratified split:
- Use n-fold stratified cross-validation 
- Use the same cross-validation folds across all models 
- For tunable models (XGBoost, Elastic Net, SVM):
  - Tune hyperparameters using cross-validation.
- Monitor performance for each damage grade, not just overall accuracy.

## Converting Model Outputs to Class Assignments

- For models that return predicted probabilities (Multinomial Logistic Regression, Elastic Net, QDA, XGBoost) for all three damage grades we will use the class with the highest probability
- If cross-validation shows poor results of a particular class look at class-specific thresholds

## Model Evaluation

- Evaluate every model using quadratic weighted kappa 
- Also show:
  - F1
  - Overall accuracy
  - Per-class precision
  - Per-class recall/sensitivity
  - Confusion matrix
- Compare both the average cross-validation performance and the consistency of performance across folds.

## Final Model Selection

- Compare all models using their cross-validated performance on the same folds.
- Use the quadratic weighted kappa to select the final model.
- Use other metrics to ensure the model is not performing well by only predicting the majority class 