---
title: "Shaw EDA"
author: "Tom Shaw and Gemini"
format: pdf
---

# Data Cleaning and Exploratory Data Analysis

## Setup and Data Loading

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, autodep = TRUE,
  fig.align = "center")
```

``` {r libraries, message=FALSE, warning=FALSE}

library(tidyverse)
library(tidymodels)
library(ggcorrplot)
library(GGally)
library(patchwork)
library(here)        # ensure filepath accuracy across all collaborator machines
```

We load the CSV files containing the training feature values and the training labels and join them together for modeling.

```{r join_tables}

# Read both CSV files
values_raw <- read_csv(
  here("data", "raw", "train_values.csv"),
  show_col_types = FALSE
)

labels_raw <- read_csv(
  here("data", "raw", "train_labels.csv"),
  show_col_types = FALSE
)

# Join as tibble
data <- values_raw |>
  inner_join(labels_raw, by = "building_id", relationship = "one-to-one") |>
  as_tibble()

# Throws an error if any rows are dropped
stopifnot(nrow(data) == nrow(values_raw))

# Save to "data/processed" for further processing and modeling
saveRDS(training_data, here("data", "processed", "earthquake.rds"))
```



## EDA and Preprocessing

First, we check for NA values across the dataset:

```{r check_NA}
c("Total NAs:" = sum(is.na(data)))
```

There are 0 NA values. Next we change the `damage_grade` outcome variable to a factor with levels "high", "medium", and "low" to improve interpretability. Our outcome of interest is "high"; we are trying to target the most damaged areas to maximize the efficient deployment of reconstruction and mitigation resources.

```{r outcome_to_factor}
data <- data |>
  mutate(
    damage_grade = factor(
      damage_grade, 
      levels = c(3, 2, 1), 
      labels = c("high", "medium", "low")
    )
  )
```

According to the [data dictionary](https://www.drivendata.org/competitions/57/nepal-earthquake/page/136/#features_list), the categorical variables and their possible values are:

- `land_surface_condition` - values n, o, t
- `foundation_type` - h, i, r, u, w
- `roof_type` - n, q, x
- `ground_floor_type` - f, m, v, x, z
- `other_floor_type` - j, q, s, x
- `position` - j, o, s, t
- `plan_configuration` - a, c, d, f, m, n, o, q, s, u
- `legal_ownership_status` - a, r, v, w

We cast all of these variables as factors:

``` {r cats_to_factors}
cat_cols <- c(
  "land_surface_condition",
  "foundation_type",
  "roof_type",
  "ground_floor_type",
  "other_floor_type",
  "position",
  "plan_configuration",
  "legal_ownership_status"
)

data <- data |>
    mutate(
      across(all_of(cat_cols), as.factor)
    ) |>
    glimpse()
```



``` {r outcome_distribution}
data |>
  ggplot(aes(x = damage_grade)) +
    geom_bar(color = "black", fill = "steelblue") +
    labs(x = "Damage Grade", y = "Frequency", title = "Outcome Distribution")

data |>
  count(damage_grade) |>
  mutate(proportion = n / sum(n))
```

The distribution of the outcome is very imbalanced with >50% of the observations in the medium category and <10% in the low category. We will need to ensure careful stratification during data splitting for validation. Next we look at the distribution for categorical variables.

```{r cat_boxplots}
df <- data |>
  select(damage_grade, where(~ is.numeric(.x) && n_distinct(na.omit(.x)) > 2)) |>
  select(-any_of(c("building_id", "geo_level_1_id", "geo_level_2_id", "geo_level_3_id"))) |>
  pivot_longer(cols = -damage_grade, names_to = "variable", values_to = "value")

ggplot(df, aes(x = damage_grade, y = value, fill = damage_grade, group = damage_grade)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  scale_fill_manual(values = c("low" = "yellow", "medium" = "orange", "high" = "red"))
```

